Grove: A Convergent Collaborative Structure-Editor Calculus

### 2021-02-10 grove-paper.tex
updates from last Grove meeting
updates with Cyrus

% indexes
\def\Root{\mathsf{Root}}
\def\LamParam{\mathsf{Lam\_param}}
\def\LamType{\mathsf{Lam\_type}}
\def\LamBody{\mathsf{Lam\_body}}
\def\AppFun{\mathsf{App\_fun}}
\def\AppArg{\mathsf{App\_arg}}
\def\PlusLeft{\mathsf{Plus\_left}}
\def\PlusRight{\mathsf{Plus\_right}}
\def\ArrowArg{\mathsf{Arrow\_arg}}
\def\ArrowResult{\mathsf{Arrow\_result}}

@@

@@

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{ll}
      \arity : \K \hfil \rightarrow & \wp(\P \times \{Exp, Pat, Typ\}) \\
      \hline
      \arity(\Root)={} & \left\{ \Root \right\} \\
      \arity(\PatVar(x))={} & \left\{ \right\} \\
      \arity(\ExpVar(x))={} & \left\{ \right\} \\
      \arity(\ExpLam)={} & \left\{ (\mathtt{Param}, Pat), (\mathtt{Type}, Typ), (\mathtt{Body}, Exp) \right\} \\
      \arity(\ExpApp)={} & \left\{ (\mathtt{Fun}, Exp), (\mathtt{Arg}, Exp) \right\} \\
      \arity(\ExpPlus)={} & \left\{ (\mathtt{L}, Exp), (\mathtt{R}, Exp) \right\} \\
      \arity(\ExpTimes)={} & \left\{ (\mathtt{L}, Exp), (\mathtt{R}, Exp) \right\} \\
      \arity(\ExpNum(n))={} & \left\{ \right\} \\
      \arity(\TypArrow)={} & \left\{ (\mathtt{Arg}, Typ), (\mathtt{Result}, Typ) \right\} \\
      \arity(\TypNum)={} & \left\{ \right\} \\
    \end{array}
  \]
  \caption{Constructors, Indexes and Arity}
  \label{fig:Constructors, Indexes and Arity}
\end{figure}

@@

Trees are often easier to work with than arbitrary graphs directly, for example
to present the graph to users or for type checking. A graph is not necessarily a
tree, but we can decompose any graph into a collection of expressions, governed
by the grammar in \autoref{fig:Base syntax}, which we organize into a structure
we call a grove.

\newvertex{DecompPlusA}{+}
\newedge{DecompTimesA}{L}
\newvertex{DecompTimesA}{*}
\newedge{DecompPlusAX}{R}
\newvertex{DecompX}{x}
\newedge{DecompTimesAX}{R}
%
\newedge{DecompPlusB}{L}
\newvertex{DecompPlusB}{+}
\newedge{DecompTimesB}{R}
\newvertex{DecompTimesB}{*}
\newedge{DecompY}{R}
\newvertex{DecompY}{y}
\newedge{DecompZ}{R}
\newvertex{DecompZ}{z}
%
\newedge{DecompTimes}{L}
\begin{figure}
  \centering
  \begin{tikzpicture}
    \path (0cm,0cm) graph[graph style] {
      root[root style] -> {
       a/{\vDecompPlusA} [> "\eSimpleTimes"] -> {
        "\vDecompTimesA" [> "\eDecompTimesA"'] -> {
         {},
         b/"\vDecompX" [> "\eDecompTimesAX"']
        },
        {}
       }
      },
    };
    \path [draw,->] (a) to ["\eDecompPlusAX",out=-60,in=60] (b);
    %
    \path (2cm,-0.85cm) graph[graph style] { plus/"\vDecompPlusB" };
    \path (3cm,-0.85cm) graph[graph style] { times/"\vDecompTimesB" };
    \path (3.25cm,-1.75cm) graph[graph style] { y/"\vDecompY" };
    \path[draw,->] (times) to ["\eDecompPlusB"'anchor=north,out=-135,in=-45] (plus);
    \path[draw,->] (plus) to ["\eDecompTimesB"anchor=south,out=45,in=135] (times);
    \path[draw,->] (times) to ["\eDecompY"anchor=west] (y);
    %
    \path (1.55cm,-2.1cm) graph[graph style] { plus/"\vDecompZ" };
  \end{tikzpicture}
  \caption{Caption goes here}%
  \label{fig:Decomposition example}
\end{figure}

@@

Let us now examine each of the four components in more detail.
The first component $e_r$ is the expression derived from the root vertex $v_0$.
\autoref{fig:Decomposition expr} defines $\expr(v)$ wich determines the expression derived from $v$.
There's one rule for each constructor form, e.g., $\ExpApp$.
In each child position, e.g., $\AppFun, \AppArg$, we call an auxiliary function $\expr'(v, p)$ which ... (explain different cases and what's going on here).

@@

% TODO: replace # with \multiVertex and \cycleVertex
% TODO: update \multiChild with more than 2 children and use it for conflict holes
% TODO: gray out ids in the grammar and expressions
% TODO: add a child to z^58 and rotate 51-52
% TODO: move example grove to right of its diagram
% TODO: distinguish between multi-parent and cycle refs

\begin{figure}
  \begin{displaymath}
    \begin{array}{l}
      \expr(v=(u, Root)) = expr'(v, Root) \\
      \expr(v=(u, \ExpVar(x))) = x^u \\
      \expr(v=(u, \ExpApp)) = (\expr'(v, \AppFun)~\expr'(v, \AppArg))^u \\
      \expr(v=(u, \ExpLam)) = \lambda^u \patt'(v, \LamParam) : \type'(v, \LamType) . \expr'(v, \LamBody) \\
    \end{array}
  \end{displaymath}

  \begin{align*}
    \expr'(v,p) = \begin{cases}
      \hole & \children(v,p) = \varnothing \\
      \expr(v') & \children(v,p) = \{v'\} \land |\parents(v')| = 1 \\
                & \phantom{\children(v,p)} \land v' \ne \min(\ancestors(v')) \\
      \#u & \children(v,p) = \{v'\} \land |\parents(v')| > 1 \\
      \#u & \children(v,p) = \{v'=(u,k)\} \land |\parents(v')| = 1 \\
          & \phantom{\children(v,p)} \land v' = \min(\ancestors(v')) \\
      \{\expr(v_1) | \ldots | \expr(v_n)\} & \children(v,p) = \{v_1, \ldots, v_n\} \\
    \end{cases}
  \end{align*}

  \begin{align*}
    \parents(v) &= \{v' : \exists \e = (u, v', p, v), G(\e) = \Plus\} \\
    \ancestors(v) &= \parents(v) \cup \ancestors(\parents(v)) \\
    \children(v, p) &= \{v' : \exists \e (u, v, p, v'), G(\e) = \Plus\} \\
    \min(\{v_1 = (u_1, k_1), \ldots, v_n = (u_n, k_n)\}) &= (u_j, k_j) \text{ s.t. } j \in \{1, \ldots, n\} \land u_j \le u_i \forall i \in \{1, \ldots, n\}
      \end{align*}
  \caption{Caption2 goes here. Note that $\ancestors(v)$ is defined as the least fixed point of the equation above.}
  \label{fig:Decomposition expr}
\end{figure}

@@

% \begin{align*}
%   \expr(v=(u, Root)) &= expr'(v, Root) \\
%   \expr(v=(u, \ExpVar(x))) &= x^u \\
%   \expr(v=(u, \ExpApp)) &= (\expr'(v, \AppFun)~\expr'(v, \AppArg))^u \\
%   % | Exp_var of (string)
%   % | Exp_lam
%   % | Exp_app
%   % | Exp_num of (int)
%   % | Exp_plus
%   % | Exp_times
%   % | Exp_nil
%   % | Exp_cons
%   % | Exp_case
%   % (**** Pat ****)
%   % | Pat_var of (string)
%   % (**** Typ ****)
%   % | Typ_num
%   % | Typ_arrow
%   % | Typ_list
% \end{align*}

>>>>

% indexes
\def\Root{\mathsf{Root}}
\def\LamParam{\mathsf{Lam\_param}}
\def\LamType{\mathsf{Lam\_type}}
\def\LamBody{\mathsf{Lam\_body}}
\def\AppFun{\mathsf{App\_fun}}
\def\AppArg{\mathsf{App\_arg}}
\def\PlusLeft{\mathsf{Plus\_left}}
\def\PlusRight{\mathsf{Plus\_right}}
\def\TimesLeft{\mathsf{Times\_left}}
\def\TimesRight{\mathsf{Times\_right}}
\def\ArrowArg{\mathsf{Arrow\_arg}}
\def\ArrowResult{\mathsf{Arrow\_result}}

@@

\newvertex{DecompPlusA}{+}
\newedge{DecompTimesA}{L}
\newvertex{DecompTimesA}{*}
\newedge{DecompPlusAX}{R}
\newvertex{DecompX}{x}
\newedge{DecompTimesAX}{R}
%
\newedge{DecompPlusB}{L}
\newvertex{DecompPlusB}{+}
\newedge{DecompTimesB}{R}
\newvertex{DecompTimesB}{*}
\newedge{DecompY}{R}
\newvertex{DecompY}{y}
\newedge{DecompZ}{R}
\newvertex{DecompZ}{z}
%
\newedge{DecompTimes}{L}
%
\newcommand{\figureDecompExample}{
\begin{figure}
  \centering
  \begin{tikzpicture}
    \path (0cm,0cm) graph[graph style] {
      root[root style] -> {
       a/{\vDecompPlusA} [> "\eSimpleTimes"] -> {
        "\vDecompTimesA" [> "\eDecompTimesA"'] -> {
         {},
         b/"\vDecompX" [> "\eDecompTimesAX"']
        },
        {}
       }
      },
    };
    \path [draw,->] (a) to ["\eDecompPlusAX",out=-60,in=60] (b);
    %
    \path (2cm,-0.85cm) graph[graph style] { plus/"\vDecompPlusB" };
    \path (3cm,-0.85cm) graph[graph style] { times/"\vDecompTimesB" };
    \path (3.25cm,-1.75cm) graph[graph style] { y/"\vDecompY" };
    \path[draw,->] (times) to ["\eDecompPlusB"'anchor=north,out=-135,in=-45] (plus);
    \path[draw,->] (plus) to ["\eDecompTimesB"anchor=south,out=45,in=135] (times);
    \path[draw,->] (times) to ["\eDecompY"anchor=west] (y);
    %
    \path (1.55cm,-2.1cm) graph[graph style] { plus/"\vDecompZ" };
  \end{tikzpicture}
  \caption{Caption goes here}%
  \label{fig:Decomposition example}
\end{figure}
}

@@

% TODO: replace positions with macros

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{ll}
      \arity : \K \hfil \rightarrow & \wp(\P \times \{Exp, Pat, Typ\}) \\
      \hline
      \arity(\Root)={} & \left\{ \Root \right\} \\
      \arity(\PatVar(x))={} & \left\{ \right\} \\
      \arity(\ExpVar(x))={} & \left\{ \right\} \\
      \arity(\ExpLam)={} & \left\{ (\LamParam, Pat), (\mathtt{Type}, Typ), (\mathtt{Body}, Exp) \right\} \\
      \arity(\ExpApp)={} & \left\{ (\mathtt{Fun}, Exp), (\mathtt{Arg}, Exp) \right\} \\
      \arity(\ExpPlus)={} & \left\{ (\mathtt{L}, Exp), (\mathtt{R}, Exp) \right\} \\
      \arity(\ExpTimes)={} & \left\{ (\mathtt{L}, Exp), (\mathtt{R}, Exp) \right\} \\
      \arity(\ExpNum(n))={} & \left\{ \right\} \\
      \arity(\TypArrow)={} & \left\{ (\mathtt{Arg}, Typ), (\mathtt{Result}, Typ) \right\} \\
      \arity(\TypNum)={} & \left\{ \right\} \\
    \end{array}
  \]
  \caption{Constructors, Indexes and Arity}
  \label{fig:Constructors, Indexes and Arity}
\end{figure}

@@

Trees are often easier to work with than arbitrary graphs directly, for example to present the graph to users or for type checking.
A graph can be decomposed into a collection of expressions governed by the grammar in \autoref{fig:Base syntax}, which we organize into a structure called a grove.

\figureDecompExample

@@

Let us now examine each of the four components in more detail.
The first component $e_r$ is the expression derived from the root vertex $v_0$.
\autoref{fig:Decomposition expr} defines $\expr(v)$ wich determines the expression derived from $v$.
There is one rule for each constructor form.
For example, the expression derived from $v_{38} = (38, \ExpPlus)$ is $\expr'(v_{38}, \PlusLeft) \text{+}^{38} \expr'(v_{38}, \PlusRight)$.
In each child position, $\expr'(v, p)$ determines the sub-expression derived from the children of $v$ at position $p$.
An empty hole $\hole$ appears if there are no children at the given position.
A conflict hole $...$  appears if there are multiple children.
If there is one child, the expression is determined recursively unless that child appears in $MP$ or $W$, discussed below, in which case we leave a reference, $\multiVertex{u}$ or $\cycleVertex{u}$ respectively.
For the $\ExpLam$ case, we define $\patt(q)$ and $\type(\tau)$ similarly.

@@

% TODO: replace # with \multiVertex and \cycleVertex
% TODO: distinguish between multi-parent and cycle refs in text
% TODO: update \multiChild with more than 2 children and use it for conflict holes
% TODO: gray out ids in the grammar and expressions
% TODO: add a child to z^58 and rotate 51-52
% TODO: move example grove to right of its diagram
% TODO: use mathord for spaces around + / *
% TODO: figure out why + --> "" and * --> "\times"
% TODO: keep condensed "|"s for conflict holes in grammar, but not other places
% TODO: make lfp a math function

\begin{figure}
  \begin{displaymath}
    \begin{array}{l}
      \expr(v=(u, Root)) = expr'(v, Root) \\
      \expr(v=(u, \ExpVar(x))) = x^u \\
      \expr(v=(u, \ExpLam)) = \lambda^u \patt'(v, \LamParam) : \type'(v, \LamType) . \expr'(v, \LamBody) \\
      \expr(v=(u, \ExpApp)) = (\expr'(v, \AppFun)~\expr'(v, \AppArg))^u \\
      \expr(v=(u, \ExpNum(n))) = n^u \\
      \expr(v=(u, \ExpPlus)) = \expr'(v, \PlusLeft)~\texttt{+}^u~\expr'(v, \PlusRight) \\
      \expr(v=(u, \ExpTimes)) = \expr'(v, \TimesLeft)~\texttt{*}^u~\expr'(v, \TimesRight) \\
      \patt(v=(u, \PatVar(x))) = x^u \\
      \type(\tau=(u, \TypArrow)) = \type'(\tau, \ArrowArg) \rightarrow \type'(\tau, \ArrowResult) \\
      \type(\tau=(u, \TypNum) = Num^u \\
    \end{array}
  \end{displaymath}

% TODO: make macros for + and *
% TODO: comment these out:
%   % | Exp_nil
%   % | Exp_cons
%   % | Exp_case
%   % | Typ_list

  \begin{align*}
    \expr'(v,p) = \begin{cases}
      \hole & \children(v,p) = \varnothing \\
      \{\expr(v_1) \mid \ldots \mid \expr(v_n)\} & \children(v,p) = \{v_1, \ldots, v_n\} \\
      \multiVertex{u} & \children(v,p) = \{v'\} \land |\parents(v')| > 1 \\
      \cycleVertex{u} & \children(v,p) = \{v'=(u,k)\} \land |\parents(v')| = 1 \\
          & \phantom{\children(v,p)} \land v' = \min(\ancestors(v')) \\
      \expr(v') & \text{otherwise} \\
    \end{cases}
  \end{align*}

  \begin{align*}
    \patt'(v,p) = \begin{cases}
      \hole & \children(v,p) = \varnothing \\
      \{\patt(v_1) \mid \ldots \mid \patt(v_n)\} & \children(v,p) = \{v_1, \ldots, v_n\} \\
      \multiVertex{u} & \children(v,p) = \{v'\} \land |\parents(v')| > 1 \\
      \cycleVertex{u} & \children(v,p) = \{v'=(u,k)\} \land |\parents(v')| = 1 \\
          & \phantom{\children(v,p)} \land v' = \min(\ancestors(v')) \\
      \patt(v') & \text{otherwise} \\
    \end{cases}
  \end{align*}

  \begin{align*}
    \type'(v,p) = \begin{cases}
      \hole & \children(v,p) = \varnothing \\
      \{\type(v_1) \mid \ldots \mid \type(v_n)\} & \children(v,p) = \{v_1, \ldots, v_n\} \\
      \multiVertex{u} & \children(v,p) = \{v'\} \land |\parents(v')| > 1 \\
      \cycleVertex{u} & \children(v,p) = \{v'=(u,k)\} \land |\parents(v')| = 1 \\
          & \phantom{\children(v,p)} \land v' = \min(\ancestors(v')) \\
      \type(v') & \text{otherwise} \\
    \end{cases}
  \end{align*}

  \begin{align*}
    \parents(v) &= \{v' : \exists \e = (u, v', p, v), G(\e) = \Plus\} \\
    \ancestors(v) &= \left( \textrm{lfp}(\ancestors') \right)(v) \\
    \ancestors'(v) &= \parents(v) \cup \ancestors'(\parents(v)) \\
    \children(v, p) &= \{v' : \exists \e (u, v, p, v'), G(\e) = \Plus\} \\
    \min(\{v_1 = (u_1, k_1), \ldots, v_n = (u_n, k_n)\}) &= (u_j, k_j) \text{ s.t. } j \in \{1, \ldots, n\} \land u_j \le u_i \forall i \in \{1, \ldots, n\}
      \end{align*}
  \caption{Caption2 goes here. Note that $\ancestors(v)$ is defined as the least fixed point of $a$.}
  \label{fig:Decomposition expr}
\end{figure}

@@
